---
title: "Transposable Element and Differential Gene Expression Analysis in Parkinson's Disease"
subtitle: "PPMI Cohort – Baseline Whole Blood RNA-seq"
author: Elizabeth Banda-Arnold
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This script reproduces the differential expression analyses (genes and HERVs/TEs)
reported in the manuscript. All data originate from the
[PPMI study](https://www.ppmi-info.org/) (Parkinson's Progression Markers
Initiative). Raw count files and clinical metadata are not redistributed here;
see **Data Availability** in the manuscript for access instructions.

### Input files expected (place in `data/` subfolder or update paths below)

| Object | Source | File |
|---|---|---|
| PPMI curated data cut | PPMI data portal | `PPMI_Curated_Data_Cut_Public_20251112.csv` |
| RNA-seq IR3 metadata | PPMI data portal | `metaDataIR3.csv` |
| Phase 1 BAM list | generated during alignment | `phase1_star_bams.txt` |
| Per-sample gene count files | featureCounts | `data/gene_counts/*.txt` |
| Per-sample HERV count files | Telescope | `data/herv_counts/*.tsv` |
| CIBERSORTx output merged with clinical | CIBERSORTx + manual merge | `data/ciber_curated_merged.csv` |
| Telescope TE annotation | Telescope v2 | `data/TE_annotation.v2.0.tsv` |

---

## 1 · Load packages

```{r packages}
suppressPackageStartupMessages({
  library(DESeq2)
  library(dplyr)
  library(readr)
  library(tidyverse)
  library(ggplot2)
  library(pheatmap)
  library(ggrepel)
  library(EnhancedVolcano)
  library(vroom)
  library(purrr)
  library(stringr)
  library(tableone)
  library(MASS)
  library(tibble)
  library(biomaRt)
  library(broom)
  library(betareg)
  library(tools)
  library(patchwork)
  library(scales)
  library(fgsea)
  library(AnnotationDbi)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(forcats)
  library(BiocParallel)
  register(SerialParam())
})
```

---

## 2 · Set paths

Edit the `DATA_DIR` variable below to point to your local data folder.
All other paths are constructed from it.

```{r paths}
DATA_DIR        <- "data"                                    
GENE_COUNT_DIR  <- file.path(DATA_DIR, "gene_counts")       # featureCounts .txt files
HERV_COUNT_DIR  <- file.path(DATA_DIR, "herv_counts")       # Telescope .tsv files

CURATED_CUT     <- file.path(DATA_DIR, "PPMI_Curated_Data_Cut_Public_20251112.csv")
RNA_META        <- file.path(DATA_DIR, "metaDataIR3.csv")
BAM_LIST        <- file.path(DATA_DIR, "phase1_star_bams.txt")
CIBER_MERGED    <- file.path(DATA_DIR, "ciber_curated_merged.csv")
TE_ANNOT        <- file.path(DATA_DIR, "TE_annotation.v2.0.tsv")
```

---

## 3 · Read and wrangle clinical & RNA metadata

```{r metadata}
curated_cut <- read_csv(CURATED_CUT, show_col_types = FALSE)
rna_meta    <- read_csv(RNA_META,    show_col_types = FALSE)

# Baseline-only RNA metadata: QC-passed, Phase 1 plates (1-28)
updated_rna_meta <- rna_meta %>%
  filter(CLINICAL_EVENT == "BL",
         QCflagIR3 != "fail",
         Plate %in% 1:28)

# Baseline-only curated clinical data
updated_curated <- curated_cut %>%
  filter(EVENT_ID == "BL")

# Participants with both baseline RNA (QC-passed) and baseline clinical data
final_cut <- updated_curated %>%
  semi_join(updated_rna_meta %>% distinct(PATNO), by = "PATNO")

# Parse BAM list; keep baseline visits only
passed_bams <- read_lines(BAM_LIST) %>%
  tibble(bam_path = .) %>%
  mutate(
    bam_file = basename(bam_path),
    PATNO    = as.integer(str_split(bam_file, "\\.", simplify = TRUE)[, 2]),
    VISIT    = str_split(bam_file, "\\.", simplify = TRUE)[, 3]
  ) %>%
  filter(VISIT == "BL")

pass_patnos <- updated_rna_meta %>%
  transmute(PATNO = as.integer(PATNO)) %>%
  distinct()

# BAMs for QC-passed, baseline subjects also present in curated cut
final_pass_bams <- passed_bams %>%
  inner_join(pass_patnos, by = "PATNO") %>%
  semi_join(final_cut %>% distinct(PATNO), by = "PATNO")
```

---

## 4 · Build analysis metadata (PD cases vs. healthy controls)

```{r build-metadata}
keep_patnos <- final_pass_bams %>% distinct(PATNO)

metadata_1 <- final_cut %>%
  semi_join(keep_patnos, by = "PATNO") %>%
  mutate(
    condition = case_when(
      COHORT == 2 & subgroup == "Healthy Control" ~ "Control",
      COHORT == 1 & subgroup != "Healthy Control" ~ "Case",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(condition)) %>%
  mutate(condition = factor(condition, levels = c("Control", "Case")))

metadata_1 %>%
  group_by(condition) %>%
  summarise(n = n_distinct(PATNO), .groups = "drop")
```

---

## 5 · Gene count matrix

### 5a · Build matrix from featureCounts output

> **Note**: Run this block once. The merged matrix is written to
> `data/combined_gene_counts.csv` so subsequent runs can skip directly
> to Section 5b.

```{r build-gene-matrix, eval=FALSE}
all_files  <- list.files(GENE_COUNT_DIR, pattern = "\\.txt$", full.names = TRUE)
bl_patnos  <- metadata_1$PATNO
file_ids   <- str_extract(basename(all_files), "(?<=IR3\\.)\\d+(?=\\.BL)")
sel_files  <- all_files[file_ids %in% bl_patnos]

count_list <- lapply(sel_files, function(f) {
  df <- read.delim(f, comment.char = "#")
  df <- df[, c(1, ncol(df))]
  colnames(df) <- c(
    "Geneid",
    sub(".*IR[0-9]+\\.([0-9]+)\\.BL.*", "\\1", basename(f))
  )
  df
})

count_matrix_raw <- Reduce(function(x, y) full_join(x, y, by = "Geneid"), count_list)

write_csv(count_matrix_raw, file.path(DATA_DIR, "combined_gene_counts.csv"))
message("Gene count matrix written.")
```

### 5b · Load matrix and remove Y-chromosome genes

```{r load-gene-matrix}
count_matrix <- read_csv(file.path(DATA_DIR, "combined_gene_counts.csv"),
                         show_col_types = FALSE) %>%
  rename_with(~ str_remove(.x, "^X")) %>%  # strip leading X added by some writers
  column_to_rownames("Geneid") %>%
  as.matrix()

storage.mode(count_matrix) <- "integer"

# Query Ensembl for Y-chromosome genes
rownames_nover <- sub("\\..*$", "", rownames(count_matrix))

mart <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl",
                   host = "https://www.ensembl.org")

gene_info <- getBM(
  attributes = c("ensembl_gene_id", "chromosome_name"),
  filters    = "ensembl_gene_id",
  values     = unique(rownames_nover),
  mart       = mart
)

y_genes <- unique(gene_info$ensembl_gene_id[
  gene_info$chromosome_name %in% c("Y", "chrY")
])

count_matrix_filtered <- count_matrix[!rownames_nover %in% y_genes, , drop = FALSE]

message(sprintf("Genes before Y-filter: %d | After: %d | Removed: %d",
                nrow(count_matrix), nrow(count_matrix_filtered),
                nrow(count_matrix) - nrow(count_matrix_filtered)))
```

---

## 6 · Covariate preparation

```{r covariates}
model_metadata <- metadata_1

# Impute single missing BMI value using age + sex linear model
bmi_lm <- lm(BMI ~ age_at_visit + SEX, data = model_metadata)
model_metadata <- model_metadata %>%
  mutate(BMI = if_else(is.na(BMI), predict(bmi_lm, newdata = model_metadata), BMI))

# Scale continuous covariates
model_metadata$BMI         <- as.numeric(scale(model_metadata$BMI))
model_metadata$age_at_visit <- as.numeric(scale(model_metadata$age_at_visit))

model_metadata$SEX <- factor(model_metadata$SEX, levels = c(1, 0), labels = c("M", "F"))
model_metadata$condition <- factor(model_metadata$condition, levels = c("Control", "Case"))
```

---

## 7 · Differential gene expression: all PD vs. Control

```{r deseq-gene}
dds_model <- DESeqDataSetFromMatrix(
  countData = count_matrix_filtered,
  colData   = model_metadata,
  design    = ~ age_at_visit + SEX + BMI + condition
)

dds_model <- DESeq(dds_model)
resultsNames(dds_model)

res_dds_model <- results(dds_model,
                          contrast = c("condition", "Case", "Control"),
                          alpha = 0.05)
summary(res_dds_model)

# Variance-stabilising transformation for QC PCA
vsd_gene <- vst(dds_model, blind = FALSE)
plotPCA(vsd_gene, intgroup = "condition")

# LFC shrinkage (apeglm)
res_lfc <- lfcShrink(dds_model,
                     coef = "condition_Case_vs_Control",
                     type = "apeglm")

# Annotate results
res_dds_model_df <- as.data.frame(res_dds_model)
res_dds_model_df$log2FoldChange_shrunk <- res_lfc$log2FoldChange

res_dds_model_df$diffexpressed <- "Not Significant"
res_dds_model_df$diffexpressed[!is.na(res_dds_model_df$padj) &
                                 res_dds_model_df$padj < 0.05 &
                                 res_dds_model_df$log2FoldChange_shrunk > 0.1] <- "Upregulated"
res_dds_model_df$diffexpressed[!is.na(res_dds_model_df$padj) &
                                 res_dds_model_df$padj < 0.05 &
                                 res_dds_model_df$log2FoldChange_shrunk < -0.1] <- "Downregulated"

genes_sig_effect <- res_dds_model_df %>%
  filter(!is.na(padj), padj < 0.05, abs(log2FoldChange_shrunk) > 1) %>%
  arrange(padj)
```

### 7a · Add HGNC gene symbols for labels

```{r gene-annotation}
res_dds_model_df$ensembl_id <- rownames(res_dds_model_df)

mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

gene_annot <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = unique(sub("\\..*$", "", res_dds_model_df$ensembl_id)),
  mart       = mart
) %>%
  mutate(across(everything(), as.character))

res_dds_model_df <- res_dds_model_df %>%
  mutate(
    ensembl_id       = as.character(ensembl_id),
    ensembl_id_clean = sub("\\..*$", "", ensembl_id)
  ) %>%
  left_join(gene_annot, by = c("ensembl_id_clean" = "ensembl_gene_id")) %>%
  mutate(gene_label = coalesce(na_if(hgnc_symbol, ""), ensembl_id_clean))

top_genes_labels <- res_dds_model_df %>%
  filter(padj < 0.05) %>%
  group_by(diffexpressed) %>%
  slice_min(padj, n = 20) %>%
  ungroup()
```

---

## 8 · HERV / TE count matrix

### 8a · Build matrix from Telescope output

> **Note**: Run once; output saved to `data/herv_matrix.csv`.

```{r build-herv-matrix, eval=FALSE}
tsv_files <- list.files(HERV_COUNT_DIR, pattern = "\\.tsv$", full.names = TRUE)

extract_sample_id <- function(f) str_extract(basename(f), "(?<=IR3\\.)\\d+(?=\\.BL)")

read_herv_counts <- function(file) {
  first_lines <- readLines(file, n = 50)
  hdr_idx     <- which(grepl("^transcript\\b", first_lines, ignore.case = TRUE))[1]
  skip_lines  <- if (!is.na(hdr_idx)) hdr_idx - 1 else 1

  df  <- read_tsv(file, skip = skip_lines, col_types = cols(.default = "c"))
  cn  <- colnames(df)
  tcol <- cn[grepl("^transcript$|locus|feature|geneid", cn, ignore.case = TRUE)][1]
  ccol <- cn[grepl("final_count|count|unique_count",    cn, ignore.case = TRUE)][1]

  if (is.na(tcol) || is.na(ccol))
    stop(paste("Missing transcript or count column in", basename(file)))

  df        <- df[, c(tcol, ccol)]
  df[[2]]   <- as.numeric(df[[2]])
  colnames(df) <- c("transcript", extract_sample_id(file))
  df
}

herv_counts_list <- lapply(tsv_files, read_herv_counts)
herv_matrix_raw  <- purrr::reduce(herv_counts_list, full_join, by = "transcript")

# Remove featureCounts summary rows
herv_matrix_raw <- herv_matrix_raw[!grepl("^__", herv_matrix_raw$transcript), ]

write_csv(herv_matrix_raw, file.path(DATA_DIR, "herv_matrix.csv"))
message("HERV matrix written.")
```

### 8b · Load and filter HERV matrix

```{r load-herv-matrix}
herv_matrix <- read_csv(file.path(DATA_DIR, "herv_matrix.csv"),
                        show_col_types = FALSE)

# Remove Y-chromosome HERV loci
herv_matrix <- herv_matrix[!grepl("_Yq|Yp", herv_matrix$transcript, ignore.case = TRUE), ]
message(sprintf("HERV loci after Y-filter: %d", nrow(herv_matrix)))

# Align metadata to matrix columns
rownames(model_metadata) <- as.character(model_metadata$PATNO)

herv_metadata_df <- model_metadata %>%
  filter(PATNO %in% colnames(herv_matrix)) %>%
  arrange(match(PATNO, colnames(herv_matrix)))

# Subset matrix to matched participants
filtered_matrix <- herv_matrix[, c("transcript", as.character(herv_metadata_df$PATNO))]
filtered_matrix[is.na(filtered_matrix)] <- 0
filtered_matrix[, -1] <- lapply(filtered_matrix[, -1], as.numeric)

herv_counts_only <- as.matrix(filtered_matrix[, -1])
rownames(herv_counts_only) <- filtered_matrix$transcript

stopifnot(all(colnames(herv_counts_only) %in% rownames(herv_metadata_df)))
message(sprintf("HERV matrix: %d loci x %d samples", nrow(herv_counts_only), ncol(herv_counts_only)))
```

---

## 9 · Differential HERV expression: all PD vs. Control

```{r deseq-herv}
dds_herv <- DESeqDataSetFromMatrix(
  countData = herv_counts_only,
  colData   = herv_metadata_df,
  design    = ~ age_at_visit + SEX + BMI + condition
)

dds_herv <- DESeq(dds_herv)
resultsNames(dds_herv)

res_dds_herv <- results(dds_herv,
                         contrast = c("condition", "Case", "Control"),
                         alpha = 0.05)
summary(res_dds_herv)

vsd_herv <- vst(dds_herv, blind = FALSE)
plotPCA(vsd_herv, intgroup = "condition")

dds_herv_lfc <- lfcShrink(dds_herv,
                           coef = "condition_Case_vs_Control",
                           type = "apeglm")

res_dds_herv_df <- as.data.frame(res_dds_herv)
res_dds_herv_df$log2FoldChange_shrunk <- dds_herv_lfc$log2FoldChange

res_dds_herv_df$diffexpressed <- "Not Significant"
res_dds_herv_df$diffexpressed[!is.na(res_dds_herv_df$padj) &
                                res_dds_herv_df$padj < 0.05 &
                                res_dds_herv_df$log2FoldChange_shrunk > 0] <- "Upregulated"
res_dds_herv_df$diffexpressed[!is.na(res_dds_herv_df$padj) &
                                res_dds_herv_df$padj < 0.05 &
                                res_dds_herv_df$log2FoldChange_shrunk < 0] <- "Downregulated"

sig_hervs <- res_dds_herv_df %>%
  filter(!is.na(padj), padj < 0.05, abs(log2FoldChange_shrunk) > 0) %>%
  arrange(padj)
```

---

## 10 · Differential HERV expression: sporadic PD (iPD) vs. Control

```{r deseq-ipd}
model4_metadata <- herv_metadata_df %>%
  filter(
    (COHORT == 1 & subgroup == "Sporadic PD") |
    (COHORT == 2 & subgroup == "Healthy Control")
  ) %>%
  mutate(
    genetic_group = case_when(
      subgroup == "Sporadic PD"    ~ "Sporadic",
      subgroup == "Healthy Control" ~ "Control"
    ),
    genetic_group = factor(genetic_group, levels = c("Control", "Sporadic"))
  )

model4_metadata$PATNO <- as.character(model4_metadata$PATNO)
counts_model4 <- herv_counts_only[, model4_metadata$PATNO, drop = FALSE]

model4_dds_herv <- DESeqDataSetFromMatrix(
  countData = counts_model4,
  colData   = model4_metadata,
  design    = ~ age_at_visit + SEX + BMI + genetic_group
)

model4_dds_herv <- DESeq(model4_dds_herv)
resultsNames(model4_dds_herv)

res_model4_dds_herv <- results(model4_dds_herv,
                               contrast = c("genetic_group", "Sporadic", "Control"),
                               alpha = 0.05)
summary(res_model4_dds_herv)

df4_dds_herv_lfc <- lfcShrink(model4_dds_herv,
                               coef = "genetic_group_Sporadic_vs_Control",
                               type = "apeglm")

res_model4_dds_herv_df <- as.data.frame(res_model4_dds_herv)
res_model4_dds_herv_df$log2FoldChange_shrunk <- df4_dds_herv_lfc$log2FoldChange

res_model4_dds_herv_df$diffexpressed <- "Not Significant"
res_model4_dds_herv_df$diffexpressed[!is.na(res_model4_dds_herv_df$padj) &
                                       res_model4_dds_herv_df$padj < 0.05 &
                                       res_model4_dds_herv_df$log2FoldChange_shrunk > 0] <- "Upregulated"
res_model4_dds_herv_df$diffexpressed[!is.na(res_model4_dds_herv_df$padj) &
                                       res_model4_dds_herv_df$padj < 0.05 &
                                       res_model4_dds_herv_df$log2FoldChange_shrunk < 0] <- "Downregulated"

sig_hervs_ipd <- res_model4_dds_herv_df %>%
  filter(!is.na(padj), padj < 0.05, abs(log2FoldChange_shrunk) > 0) %>%
  arrange(padj)
```

---

## 11 · Differential HERV expression: LRRK2 carriers vs. Control

> This block requires `model3_dds_herv`, a DESeq2 object built from a
> metadata subset containing LRRK2 carriers and healthy controls
> (construction follows the same pattern as Section 10).

```{r deseq-lrrk2}
# -- Build model3_dds_herv for LRRK2 carriers + Controls (mirror Section 10) --
model3_metadata <- herv_metadata_df %>%
  filter(
    (COHORT == 1 & subgroup == "LRRK2 PD") |
    (COHORT == 2 & subgroup == "Healthy Control")
  ) %>%
  mutate(
    genetic_group = case_when(
      subgroup == "LRRK2 PD"       ~ "LRRK2_carrier",
      subgroup == "Healthy Control" ~ "Control"
    ),
    genetic_group = factor(genetic_group, levels = c("Control", "LRRK2_carrier"))
  )

model3_metadata$PATNO <- as.character(model3_metadata$PATNO)
counts_model3 <- herv_counts_only[, model3_metadata$PATNO, drop = FALSE]

model3_dds_herv <- DESeqDataSetFromMatrix(
  countData = counts_model3,
  colData   = model3_metadata,
  design    = ~ age_at_visit + SEX + BMI + genetic_group
)
model3_dds_herv <- DESeq(model3_dds_herv)

# -- Contrast: LRRK2 carrier vs. Control --
res_lrrk2_control <- results(model3_dds_herv,
                             contrast = c("genetic_group", "LRRK2_carrier", "Control"),
                             alpha = 0.05)
summary(res_lrrk2_control)

res_lrrk2_dds_herv_lfc <- lfcShrink(model3_dds_herv,
                                    coef = "genetic_group_LRRK2_carrier_vs_Control",
                                    type = "apeglm")

res_lrrk2_control_df <- as.data.frame(res_lrrk2_control)
res_lrrk2_control_df$log2FoldChange_shrunk <- res_lrrk2_dds_herv_lfc$log2FoldChange

res_lrrk2_control_df$diffexpressed <- "Not Significant"
res_lrrk2_control_df$diffexpressed[!is.na(res_lrrk2_control_df$padj) &
                                     res_lrrk2_control_df$padj < 0.05 &
                                     res_lrrk2_control_df$log2FoldChange_shrunk > 0] <- "Upregulated"
res_lrrk2_control_df$diffexpressed[!is.na(res_lrrk2_control_df$padj) &
                                     res_lrrk2_control_df$padj < 0.05 &
                                     res_lrrk2_control_df$log2FoldChange_shrunk < 0] <- "Downregulated"

sig_lrrk2 <- res_lrrk2_control_df %>% filter(padj < 0.05)
```

---

## 12 · CIBERSORTx deconvolution – beta regression

### 12a · Prepare CIBERSORTx mixture file

> Run once after generating normalised counts. Upload the mixture file
> to the CIBERSORTx portal; merge the results with clinical metadata
> before Section 12b.

```{r cibersortx-prep, eval=FALSE}
cibersortdata <- counts(dds_model, normalized = TRUE)
ensembl_ids   <- sub("\\..*$", "", rownames(cibersortdata))

mart <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl",
                   mirror = "useast")

gene_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = unique(ensembl_ids),
  mart       = mart
) %>%
  filter(hgnc_symbol != "")

counts_df <- as.data.frame(cibersortdata)
counts_df$ensembl_gene_id <- ensembl_ids

cibersort_df <- counts_df %>%
  inner_join(gene_map, by = "ensembl_gene_id") %>%
  dplyr::select(-ensembl_gene_id) %>%
  group_by(hgnc_symbol) %>%
  summarise(across(where(is.numeric), sum), .groups = "drop") %>%
  rename(GeneSymbol = hgnc_symbol)

write.table(cibersort_df,
            file = file.path(DATA_DIR, "cibersort_mixture.txt"),
            sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)

# Sample labels (optional, for CIBERSORTx batch correction)
labels <- metadata_1 %>%
  transmute(SampleID = PATNO,
            Condition = if_else(subgroup == "Healthy Control", "HC", "PD")) %>%
  distinct()

write.table(labels,
            file = file.path(DATA_DIR, "cibersortx_labels.txt"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

### 12b · Beta regression on CIBERSORTx cell fractions

```{r beta-regression}
dat <- read_csv(CIBER_MERGED, show_col_types = FALSE)

analysis_dat <- dat %>%
  mutate(
    group = case_when(COHORT == 2 ~ "HC", COHORT == 1 ~ "PD", TRUE ~ NA_character_),
    group = factor(group, levels = c("HC", "PD")),
    SEX   = factor(SEX)
  ) %>%
  filter(!is.na(group), !is.na(P.value), P.value < 0.05)  # retain well-deconvolved samples

cell_types <- c(
  "Neutrophils",
  "Monocytes",
  "NK.cells.resting",
  "T.cells.CD4.memory.resting",
  "T.cells.CD4.memory.activated",
  "T.cells.CD4.naive",
  "B.cells.naive"
)

missing_cells <- setdiff(cell_types, names(analysis_dat))
if (length(missing_cells) > 0)
  stop("Missing cell-type columns: ", paste(missing_cells, collapse = ", "))

long_dat <- analysis_dat %>%
  select(PATNO, group, age_at_visit, SEX, BMI, all_of(cell_types)) %>%
  pivot_longer(cols = all_of(cell_types),
               names_to  = "cell_type",
               values_to = "cell_fraction") %>%
  filter(across(c(cell_fraction, age_at_visit, SEX, BMI), ~ !is.na(.)))

# Beta regression (Smithson-Verkuilen transformation)
fit_beta_model <- function(df_cell) {
  n <- nrow(df_cell)
  df_cell <- df_cell %>%
    mutate(cell_fraction_beta = (cell_fraction * (n - 1) + 0.5) / n)

  model <- betareg(cell_fraction_beta ~ group + age_at_visit + SEX + BMI,
                   data = df_cell, link = "logit")

  broom::tidy(model, conf.int = TRUE) %>%
    filter(component == "mean", term == "groupPD") %>%
    mutate(n = n, mean_cell_fraction = mean(df_cell$cell_fraction, na.rm = TRUE))
}

# Linear model sensitivity analysis
fit_lm_model <- function(df_cell) {
  model <- lm(cell_fraction ~ group + age_at_visit + SEX + BMI, data = df_cell)
  broom::tidy(model, conf.int = TRUE) %>%
    filter(term == "groupPD") %>%
    mutate(n = nrow(df_cell), mean_cell_fraction = mean(df_cell$cell_fraction, na.rm = TRUE))
}

beta_results <- long_dat %>%
  group_by(cell_type) %>%
  group_split() %>%
  set_names(map_chr(., ~ unique(.x$cell_type))) %>%
  map_dfr(fit_beta_model, .id = "cell_type") %>%
  mutate(p_adj_bh = p.adjust(p.value, method = "BH"),
         significant_fdr = p_adj_bh < 0.05) %>%
  arrange(p_adj_bh)

lm_results <- long_dat %>%
  group_by(cell_type) %>%
  group_split() %>%
  set_names(map_chr(., ~ unique(.x$cell_type))) %>%
  map_dfr(fit_lm_model, .id = "cell_type") %>%
  mutate(lm_p_adj_bh  = p.adjust(p.value, method = "BH"),
         lm_significant_fdr = lm_p_adj_bh < 0.05) %>%
  rename(lm_estimate = estimate, lm_std_error = std.error,
         lm_statistic = statistic, lm_p_value = p.value,
         lm_conf_low = conf.low, lm_conf_high = conf.high) %>%
  arrange(lm_p_adj_bh)

final_results <- beta_results %>%
  rename(beta_estimate = estimate, beta_std_error = std.error,
         beta_statistic = statistic, beta_p_value = p.value,
         beta_conf_low = conf.low, beta_conf_high = conf.high) %>%
  left_join(lm_results %>% select(cell_type, starts_with("lm_")), by = "cell_type") %>%
  arrange(p_adj_bh)

print(final_results)
```

---

## 13 · HERV family enrichment

```{r herv-family}
herv_annot <- read_tsv(TE_ANNOT, show_col_types = FALSE)

# Significant DE HERVs from the all-PD model (Section 9)
sig_dds_herv_df <- res_dds_herv_df %>%
  filter(!is.na(padj), padj < 0.05)

# Remove L1 elements (separate TE class from HERVs)
herv_annot_clean <- herv_annot %>% filter(Family != "L1")

sig_hervs_freq_clean <- sig_dds_herv_df %>%
  rownames_to_column("Locus") %>%
  inner_join(herv_annot_clean %>% select(Locus, Family), by = "Locus")

herv_family_counts_db <- herv_annot_clean %>% count(Family, name = "db_n")
herv_family_counts_de <- sig_hervs_freq_clean %>% count(Family, name = "de_n")

family_counts <- herv_family_counts_db %>%
  full_join(herv_family_counts_de, by = "Family") %>%
  mutate(
    db_n = coalesce(db_n, 0L),
    de_n = coalesce(de_n, 0L),
    log10_db      = ifelse(db_n > 0, log10(db_n), NA_real_),
    log10_de      = ifelse(de_n > 0, log10(de_n), NA_real_),
    ratio_de_over_db = ifelse(db_n > 0, de_n / db_n, NA_real_)
  ) %>%
  arrange(desc(db_n)) %>%
  mutate(Family = factor(Family, levels = Family))

expected_line <- sum(family_counts$de_n) / sum(family_counts$db_n)

family_counts <- family_counts %>%
  mutate(fold_enrichment = ratio_de_over_db / expected_line)

pA <- ggplot(family_counts, aes(x = Family, y = log10_db)) +
  geom_col() +
  labs(x = "HERV Family", y = "log10 frequency (database)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

pB <- ggplot(family_counts, aes(x = Family, y = log10_de)) +
  geom_col(fill = "red3") +
  labs(x = "HERV Family", y = "log10 frequency (DE loci)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

pC <- ggplot(family_counts,
             aes(x = Family, y = ratio_de_over_db,
                 fill = ratio_de_over_db > expected_line)) +
  geom_col() +
  scale_fill_manual(values = c("gray70", "deepskyblue3"), guide = "none") +
  geom_hline(yintercept = expected_line, linetype = "dashed", color = "red") +
  labs(x = "HERV Family", y = "DE / database ratio") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

(pA / pB / pC) +
  plot_annotation(tag_levels = "A",
                  title = "HERV Family Frequencies (L1 removed)")
```

---

## 14 · Proximal gene set enrichment (GSEA)

```{r gsea}
herv_annot <- read_tsv(TE_ANNOT, show_col_types = FALSE)

all_hervs_df <- as.data.frame(res_dds_herv_df) %>%
  rownames_to_column("Locus") %>%
  select(Locus, log2FoldChange_shrunk, padj, stat, baseMean)

herv_gene_long <- all_hervs_df %>%
  left_join(herv_annot, by = "Locus") %>%
  transmute(
    Locus,
    score = sign(log2FoldChange_shrunk) * -log10(padj),
    ClosestUpstream_id,
    ClosestDownstream_id,
    IntersectedGeneID
  ) %>%
  pivot_longer(
    cols = c(ClosestUpstream_id, ClosestDownstream_id, IntersectedGeneID),
    names_to  = "gene_relation",
    values_to = "gene_id"
  ) %>%
  filter(!is.na(gene_id), !is.na(score))

collapse_max_abs <- function(x) x[which.max(abs(x))]

gene_scores <- herv_gene_long %>%
  group_by(gene_id) %>%
  summarise(score = collapse_max_abs(score), .groups = "drop") %>%
  separate_rows(gene_id, sep = ",")

entrez_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys    = unique(gene_scores$gene_id),
  keytype = "ENSEMBL",
  columns = "ENTREZID"
)

gene_scores_entrez <- gene_scores %>%
  left_join(entrez_map, by = c("gene_id" = "ENSEMBL")) %>%
  filter(!is.na(ENTREZID)) %>%
  group_by(ENTREZID) %>%
  slice_max(order_by = abs(score), n = 1, with_ties = FALSE) %>%
  ungroup()

ranks <- setNames(gene_scores_entrez$score, gene_scores_entrez$ENTREZID)

# Break remaining ties with micro-jitter (ensures reproducability)
set.seed(42)
tie_mask <- duplicated(ranks) | duplicated(ranks, fromLast = TRUE)
if (any(tie_mask))
  ranks[tie_mask] <- ranks[tie_mask] + runif(sum(tie_mask), -1e-7, 1e-7)

ranks <- sort(ranks, decreasing = TRUE)

gsea_go_bp <- gseGO(
  geneList      = ranks,
  OrgDb         = org.Hs.eg.db,
  ont           = "BP",
  keyType       = "ENTREZID",
  eps           = 0,
  minGSSize     = 10,
  maxGSSize     = 800,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE
)

dotplot(gsea_go_bp, showCategory = 10, split = ".sign") +
  facet_grid(. ~ .sign)

cat("Activated BP terms:", sum(gsea_go_bp@result$NES > 0), "\n")
cat("Suppressed BP terms:", sum(gsea_go_bp@result$NES < 0), "\n")
```

---

## 15 · Manuscript figure theme

```{r figure-theme}
my_nature_theme <- theme(
  text           = element_text(size = 9),
  plot.title     = element_text(size = 11, face = "bold", hjust = 0.5,
                                margin = margin(b = 4)),
  axis.title     = element_text(size = 10, face = "bold"),
  axis.text      = element_text(size = 9, color = "black"),
  axis.line      = element_line(linewidth = 0.5),
  panel.border   = element_rect(color = "black", fill = NA, linewidth = 0.5),
  strip.text     = element_text(size = 10, face = "bold"),
  strip.background = element_rect(fill = "white", color = "black", linewidth = 0.5),
  legend.title   = element_text(size = 10, face = "bold"),
  legend.text    = element_text(size = 9),
  plot.margin    = margin(5, 5, 5, 5)
)
```

---

## Session info

```{r session-info}
sessionInfo()
```
